Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-11 appendix.tex
fidelities --> branch lengths

See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
\label{fig:bl-inconsistency}
\end{figure}

>>>>

See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < p_{x^*}, p_{y^*} < 1$ with $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Analytically-derived region of branch parameter inconsistency where $p_x=p_{x^*}$ and $p_y=p_{y^*}$ are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
The shaded region shows the area in which the internal branch length $\hat{p}_w$ is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
\label{fig:bl-inconsistency}
\end{figure}

### 2018-05-11 joint_inf.tex
fidelities --> branch lengths

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

%das: TODO update this intuition
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
    As in previous plots, the shaded and white regions are loose indications of inconsistency.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name figures/w-hat-empirical-01.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01}
\caption{
    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

>>>>

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$ ($p_{w^*}=p_{y^*}$).
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
    In the marked region, either $\hat{p}_w=0$ or, for $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$, one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
    As in previous plots, the shaded and white regions are loose indications of inconsistency.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name ../intermed/w-hat-empirical-01.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
}
\label{fig:bl-general-inconsistency}
\end{figure}
